---
title: "How do TIFs impact municipal tax bases?"
format:
  html:
    embed-resources: true
editor: visual
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(glue)
library(here)
library(DBI)
library(RSQLite)
library(ptaxsim)
library(vctrs)
library(dplyr)
library(tidyverse)
library(httr)
library(jsonlite)
library(ccao)
library(ggplot2)
library(plotly)
library(statip)
library(scales)
library(crosstalk)
library(htmltools)
library(sf)
library(ggspatial)
library(prettymapr)
library(purrr)
library(RColorBrewer)
```

# Introduction

Tax increment financing (TIF) is a tool used to promote economic development in specific areas. It allows a municipality to re-invest all incremental property tax revenue back into the area from which it came. TIF districts last for 23 years, with the option to extend to 35 years.

TIFs work by "freezing" the value of all property within them at the time they are established. Any "new" property value created within the TIF (whether through development or rising prices) is taxed as usual. However, the tax revenue generated by any new value over the frozen amount is diverted into the TIF fund, which is then allocated for specific projects. State law mandates that the amount of TIF surplus declared by the City of Chicago is distributed among the local governments, such as CPS, based on their share of total property tax.

TIF districts are designed by the City and approved by the City Council. Some of the primary incentives municipalities have to establish TIFs are as follows:

-   **Funding Infrastructure and Capital Projects:** TIFs are used to dedicate a portion of property tax collections to fund specific redevelopment costs, such as public infrastructure, renovations, and maintenance.
-   **Earmarking Growth:** By freezing the taxable value, Equalized Assessed Value (EAV), at a "base" level, the city can earmark all future incremental growth in property value specifically for projects within that district's boundaries.
-   **Incentivizing Investment:** TIF districts are designated to encourage growth that might not otherwise occur, allowing the city to invest in its own sister agencies (like the CTA or Chicago Public Schools) through intergovernmental agreements for school renovations and other improvements.

### Why do we investigate TIFs and Municipal Tax Bases?

Understanding how TIFs impact municipal tax bases is critical because they fundamentally reshape how public revenue is distributed across taxing agencies. While TIF districts are designed to stimulate development, they do so by diverting growth in property values away from the general tax base and into a restricted fund. This creates a structural tradeoff: investment within the TIF may generate long-term economic benefits, but in the short term it reduces the taxable base available to overlapping jurisdictions such as school districts, park districts, and municipal governments. As a result, these agencies must often increase tax rates to meet their fixed revenue needs, shifting the burden onto taxpayers outside the TIF or within the frozen base. In places like [Cook County]{.underline}, where TIF usage is widespread and, in some municipalities, highly concentrated, this dynamic can significantly influence fiscal equity, public service funding, and long-term financial stability. This analysis therefore examines not just where TIFs are used, but how their design translates into measurable impacts on tax rates and the distribution of public resources over time.

# Part I: Identifying cities with high proportion of TIF-ed PINs

## Identifying cities with high proportion of PINs under TIFs.

### City of Chicago TIFed PINs

This report identifies top five cities apart from the City of Chicago, within Cook County with the highest ratio of total PINs (Proprty Identification Number) and PINs within TIF districts. For comparison, let's calculate the total PINs to TIFed PINs ratio for the city of Chicago.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Get City of Chicago PINs 
base_url <- "https://datacatalog.cookcountyil.gov/resource/nj4t-kc8j.json"

chicago_raw_data <- GET(
  base_url,
  query = list(
    tax_year = 2023,
    `$select` = "pin, cook_municipality_name, tax_tif_district_num",
    `$where` = "triad_code = '1'",
    `$limit` = 1000000L
  )
)

chicago_pins <- fromJSON(rawToChar(chicago_raw_data$content))
rm(chicago_raw_data)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
chicago_tifed_pins <- chicago_pins %>%
  filter(cook_municipality_name == "CITY OF CHICAGO") %>%
  group_by(cook_municipality_name) %>%
  summarise(
    total_pins = n(),
    tif_pins = sum(!is.na(tax_tif_district_num)),
    muni_tif_ratios = tif_pins / total_pins,
    .groups = "drop"
  )

chicago_tifed_pins %>%
  mutate(muni_tif_ratios = scales::percent(muni_tif_ratios, accuracy = 0.1)) %>%
  rename(
    `Municipality` = cook_municipality_name,
    `Total PINs` = total_pins,
    `TIF-ed PINs` = tif_pins,
    `TIF Ratio` = muni_tif_ratios
  ) %>%
  DT::datatable(options = list(dom = 't'), rownames = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dist_frequency <- chicago_pins %>%
  count(tax_tif_district_num)
```

Of the 882,969 PINs in Chicago, 220,942 (approximately 25%) are located within the city's 124 TIF districts. These districts are significant drivers of local finance, accounting for 17.8% of the total tax billed across all taxing districts in the city. ([Source](https://share.google/DLxKU5savfVWmcGyL))

### Cities Across Cook County

```{r, echo=FALSE, message=FALSE, warning=FALSE}
triad_2_3_raw_data <- GET(
  base_url,
  query = list(
    tax_year = 2023,
    `$select` = "pin, cook_municipality_name, tax_tif_district_num",
    `$where` = "triad_code > '1'",
    `$limit` = 1000000L
  )
)

triad_2_3_pins <- fromJSON(rawToChar(triad_2_3_raw_data$content))
rm(triad_2_3_raw_data)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
triad_2_3_tifed_ratios <- triad_2_3_pins %>%
  filter(!cook_municipality_name %in% c("CITY OF CHICAGO", "VILLAGE OF EAST DUNDEE")) %>%
  group_by(cook_municipality_name) %>%
  summarise(
    total_pins = n(),
    tif_pins = sum(!is.na(tax_tif_district_num)),
    muni_tif_ratios = tif_pins / total_pins,
    .groups = "drop"
  ) %>%
  slice_max(muni_tif_ratios, n = 5)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
triad_2_3_tifed_ratios %>%
  mutate(muni_tif_ratios = scales::percent(muni_tif_ratios, accuracy = 1)) %>%
  rename(
    `Municipality` = cook_municipality_name,
    `Total PINs` = total_pins,
    `TIF-ed PINs` = tif_pins,
    `TIF Ratio` = muni_tif_ratios
  ) %>%
  DT::datatable(
    options = list(dom = 't', paging = FALSE),
    rownames = FALSE,
    caption = 'Top 5 Municipalities by TIF-to-Total PIN Ratio (Triads 2 & 3)'
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
target_cities_raw <- GET(
  base_url,
  query = list(
    tax_year = 2023,
    `$select` = "pin, cook_municipality_name, tax_tif_district_num, tax_tif_district_name, tax_code, township_name, triad_name, triad_code",
    `$where` = "cook_municipality_name IN('VILLAGE OF PHOENIX', 'VILLAGE OF BEDFORD PARK', 'VILLAGE OF BELLWOOD', 'VILLAGE OF FORD HEIGHTS', 'VILLAGE OF POSEN')",
    `$limit` = 1000000L
  )
)

target_cities <- fromJSON(rawToChar(target_cities_raw$content))
rm(target_cities_raw)
```

The five municipalities identified are Village of Phoenix, Village of Bedford Park, Village of Bellwood, Village of Ford Heights, and Village of Posen. All are located in the south and western portions of the [Chicago metropolitan area]{.underline}. Their geographic clustering is notable, as these regions have historically relied more heavily on TIF districts as a development tool, making them particularly relevant for understanding how concentrated TIF usage shapes local tax bases and fiscal conditions.

## Visualizing TIFs and Parcel Types

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ptaxsim_db_conn <- dbConnect(SQLite(), "ptaxsim.db")

target_pin10s <- unique(substr(target_cities$pin, 1, 10))

target_pins_geo <- lookup_pin10_geometry(year = 2023, pin10 = target_pin10s)
target_pins_geo <- target_pins_geo %>%
  st_as_sf(wkt = "geometry", crs = 4326)

target_pins_class <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  glue_sql("
    SELECT pin, class
    FROM pin
    WHERE pin IN ({target_cities$pin*})
    AND year = 2023
    ",
    .con = ptaxsim_db_conn
  )
)

target_pins_geo <- target_pins_geo %>%
  left_join(
    target_pins_class %>%
      mutate(pin10 = substr(pin, 1, 10)) %>%
      group_by(pin10) %>%
      summarize(prop_class = first(class)),
    by = "pin10"
  ) %>%
  left_join(
    target_cities %>%
      mutate(pin10 = substr(pin, 1, 10)) %>%
      group_by(pin10) %>%
      summarize(
        municipality = first(cook_municipality_name),
        tif_district = first(tax_tif_district_name)
      ),
    by = "pin10"
  ) %>%
  mutate(
    major_class = substr(prop_class, 1, 1),
    major_class_fct = recode_factor(
      major_class,
      "2" = "2 - Residential",
      "3" = "3 & 5 - Commercial",
      "5" = "3 & 5 - Commercial",
      "0" = "Other", "1" = "Other", "6" = "Other", "7" = "Other"
    ),
    in_tif = !is.na(tif_district) & tif_district != ""
  )

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
read_geojson_url <- function(url) {
  tmp <- tempfile(fileext = ".geojson")
  httr::GET(url, httr::write_disk(tmp, overwrite = TRUE),
            httr::config(ssl_verifypeer = FALSE))
  st_read(tmp, quiet = TRUE)
}

bound_muni <- read_geojson_url(paste0(
  "https://opendata.arcgis.com/api/v3/datasets/",
  "534226c6b1034985aca1e14a2eb234af_2/downloads/",
  "data?format=geojson&spatialRefId=4326&where=1%3D1"
)) %>%
  filter(MUNICIPALITY %in% c(
    "Phoenix", "Bedford Park", "Bellwood",
    "Ford Heights", "Posen"
  ))

bound_tif <- read_geojson_url(paste0(
  "https://opendata.arcgis.com/api/v3/datasets/",
  "8aeb00520c544aafb9a22510465c679d_18/downloads/",
  "data?format=geojson&spatialRefId=4326&where=1%3D1"
))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tif_agencies <- target_cities %>%
  filter(!is.na(tax_tif_district_num) & tax_tif_district_num != "") %>%
  distinct(tax_tif_district_num) %>%
  pull(tax_tif_district_num)

bound_tif <- bound_tif %>%
  filter(AGENCYNUM %in% tif_agencies)
```

### Maps of the five identified cities

Use the dropdown menu to select a city of interest.

```{r echo=FALSE, message=FALSE, warning=FALSE}
map_plot        <- plot_ly()
traces_per_city <- c()
all_city_geo    <- list()

all_levels <- levels(target_pins_geo$major_class_fct)
pal <- setNames(
  colorRampPalette(brewer.pal(9, "Set1"))(length(all_levels)),
  all_levels
)

target_pins_geo <- st_cast(target_pins_geo, "MULTIPOLYGON")
bound_muni      <- st_cast(bound_muni, "MULTIPOLYGON")
bound_tif       <- st_cast(bound_tif,  "MULTIPOLYGON")

city_names <- c(
  "VILLAGE OF PHOENIX",
  "VILLAGE OF BEDFORD PARK",
  "VILLAGE OF BELLWOOD",
  "VILLAGE OF FORD HEIGHTS",
  "VILLAGE OF POSEN"
)

for (city in city_names) {
  city_pins  <- target_pins_geo %>% filter(municipality == city)
  clean      <- tools::toTitleCase(tolower(gsub("VILLAGE OF ", "", city)))
  city_muni  <- bound_muni %>% filter(MUNICIPALITY == clean)
  city_tifs  <- st_filter(bound_tif, city_muni)
  bb         <- st_bbox(city_muni)
  pad        <- 0.01

  all_city_geo[[city]] <- list(
    "geo.lonaxis.range" = list(bb[["xmin"]] - pad, bb[["xmax"]] + pad),
    "geo.lataxis.range" = list(bb[["ymin"]] - pad, bb[["ymax"]] + pad)
  )

  trace_count <- 0L

  for (cls in levels(target_pins_geo$major_class_fct)) {
    sub <- filter(city_pins, major_class_fct == cls)
    if (nrow(sub) == 0L) next
    map_plot <- map_plot %>% add_sf(
      data          = sub,
      fillcolor     = alpha(pal[[cls]], 0.5),
      stroke        = I(pal[[cls]]),
      span          = I(0.8),
      name          = cls,
      legendgroup   = cls,
      showlegend    = TRUE,
      visible       = FALSE,
      hovertemplate = "none"
    )
    trace_count <- trace_count + 1L
  }

  map_plot <- map_plot %>% add_sf(
    data        = city_muni,
    fillcolor   = "transparent",
    stroke      = I("black"),
    span        = I(2),
    name        = "Municipal Boundary",
    legendgroup = "Municipal Boundary",
    showlegend  = TRUE,
    visible     = FALSE,
    hoverinfo   = "none"
  )
  trace_count <- trace_count + 1L

  if (nrow(city_tifs) > 0L) {
    map_plot <- map_plot %>% add_sf(
      data          = city_tifs,
      fillcolor     = alpha("purple", 0.3),
      stroke        = I("purple4"),
      span          = I(1.5),
      name          = "TIF District",
      legendgroup   = "TIF District",
      showlegend    = TRUE,
      visible       = FALSE,
      hovertemplate = "none"
    )
    trace_count <- trace_count + 1L
  }

  traces_per_city <- c(traces_per_city, trace_count)
}

cumulative_traces <- c(0, cumsum(traces_per_city))
total_traces      <- sum(traces_per_city)

buttons <- lapply(seq_along(city_names), function(i) {
  vis <- rep(FALSE, total_traces)
  vis[(cumulative_traces[i] + 1):cumulative_traces[i + 1]] <- TRUE
  list(
    method = "update",
    label  = gsub("VILLAGE OF ", "", city_names[i]),
    args   = list(
      list(visible = vis),
      c(
        list(title = list(text = gsub("VILLAGE OF ", "", city_names[i]), x = 0.5)),
        all_city_geo[[city_names[i]]]
      )
    )
  )
})

init     <- all_city_geo[[city_names[1]]]
map_plot <- map_plot %>% layout(
  title = list(text = gsub("VILLAGE OF ", "", city_names[1]), x = 0.5),
  geo = list(
    showland       = TRUE,
    landcolor      = "rgb(235, 235, 235)",
    showcoastlines = FALSE,
    showframe      = FALSE,
    lonaxis        = list(range = init[["geo.lonaxis.range"]]),
    lataxis        = list(range = init[["geo.lataxis.range"]])
  ),
  legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.05),
  updatemenus = list(list(
    active      = 0,
    buttons     = buttons,
    x           = 0.01,
    xanchor     = "left",
    y           = 1.12,
    yanchor     = "top",
    bgcolor     = "white",
    bordercolor = "#aaa",
    font        = list(size = 13)
  )),
  margin = list(t = 90, b = 60)
)

map_plot <- plotly_build(map_plot)
for (j in 1:traces_per_city[1]) {
  map_plot$x$data[[j]]$visible <- TRUE
}
map_plot
```

# Part II: Reviewing Increments

This section explains how taxable values have grown within TIF districts. When a TIF is created, the Equalized Assessed Value (EAV) of all property within it is 'frozen' at a base level. As property values rise, the growth above that frozen base, known as the increment, is measured in EAV, not dollars. Tax revenue is then generated by applying the local tax rate to that increment, and the resulting revenue is funneled into the TIF fund. Meanwhile, the frozen base EAV remains available for other taxing agencies such as municipalities and school districts as usual.

### Step 1: Prepare data

Add a municipality name column to the dataset of unique TIF tax codes so that all records can be aggregated and analyzed at the city level.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Re-establish DB connection
ptaxsim_db_conn <- dbConnect(SQLite(), "ptaxsim.db")

# Query tif_distribution directly for frozen EAV — this is the correct table
tif_dists <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  glue::glue_sql("
    SELECT *
    FROM tif_distribution
    WHERE year IN ({2006:2023*})
    AND tax_code_num IN ({as.integer(unique(target_cities$tax_code))*})
  ", .con = ptaxsim_db_conn)
)

# Aggregate frozen EAV to agency level
tif_dists_agency <- tif_dists %>%
  group_by(year, agency_num) %>%
  summarize(
    agency_frozen_eav = sum(as.numeric(tax_code_frozen_eav), na.rm = TRUE),
    .groups = "drop"
  )

# Stable frozen EAV: anchor to each TIF's earliest recorded year
tif_frozen_stable <- tif_dists_agency %>%
  filter(agency_frozen_eav > 0) %>%
  group_by(agency_num) %>%
  slice_min(year, n = 1, with_ties = FALSE) %>%
  rename(tif_creation_year = year, frozen_eav_stable = agency_frozen_eav)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Get all pins within TIF tax codes and their full history
tif_pins_vec <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  glue::glue_sql("
    SELECT DISTINCT pin
    FROM pin
    WHERE tax_code_num IN ({unique(tif_dists$tax_code_num)*})
  ", .con = ptaxsim_db_conn)
) %>%
  pull(pin)

tif_pins_dt <- lookup_pin(2006:2023, pin = tif_pins_vec) %>%
  mutate(tax_code = lookup_tax_code(year, pin))
```

### Step 2: Summarize EAV

Aggregate the total Equalized Assessed Value (EAV) by year, municipality, and TIF district to calculate Total EAV. To estimate the incremental value, or Amount to TIF, subtract the district’s frozen EAV from the current year’s total EAV. In this step, a stable frozen baseline is also established by anchoring each district’s frozen EAV to its earliest recorded year. This ensures that any later administrative changes to the baseline do not affect the increment calculation.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tif_pins_summ <- tif_pins_dt %>%
  left_join(
    target_cities %>%
      select(pin, cook_municipality_name, tax_tif_district_num) %>%
      distinct(),
    by = "pin"
  ) %>%
  left_join(
    tif_dists %>%
      mutate(tax_code_num = as.character(tax_code_num)) %>%
      select(year, tax_code_num, agency_num) %>%
      distinct(),
    by = c("year", "tax_code" = "tax_code_num")
  ) %>%
  left_join(tif_frozen_stable, by = "agency_num") %>%
  group_by(year, cook_municipality_name, agency_num,
           tif_creation_year, frozen_eav_stable) %>%
  summarize(total_eav = sum(eav), .groups = "drop") %>%
  mutate(
    amt_to_tif = pmax(total_eav - frozen_eav_stable, 0)
  )
```

### Step 3: Summarize Increments

Sum the Amount to TIF across all TIF districts at the city and year level to obtain total increment values for each municipality over time.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
city_totals <- tif_pins_summ %>%
  group_by(year, cook_municipality_name) %>%
  summarize(
    total_eav  = sum(total_eav, na.rm = TRUE),
    frozen_eav = sum(frozen_eav_stable, na.rm = TRUE),
    amt_to_tif = sum(amt_to_tif, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(frozen_eav = if_else(frozen_eav == 0, NA_real_, frozen_eav))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tif_creation_by_city <- tif_pins_summ %>%
  select(cook_municipality_name, agency_num, tif_creation_year) %>%
  distinct() %>%
  filter(!is.na(tif_creation_year)) %>%
  left_join(
    target_cities %>%
      select(tax_tif_district_num, tax_tif_district_name) %>%
      distinct(),
    by = c("agency_num" = "tax_tif_district_num")
  )
```

### Step 4: Visualizing Increments

Generate a time-series visualization for each municipality to track the growth of TIF increments relative to the frozen base. This process includes filtering the pre-aggregated city-level dataset for a selected municipality, applying the stable frozen baseline as a consistent reference point, calculating the increment as the difference between Total EAV and the baseline while restricting values to be non-negative, and annotating the chart to indicate the creation year of each TIF district.

The charts below visualize the growth of TIF increments relative to the frozen base for each municipality over time. Dahsed verticle lines represent introduction of a new TIF district. Use the dropdown menu to select a city and explore how these patterns differ across locations.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
muni_plot           <- plot_ly()
traces_per_city_v2  <- c()
all_city_annotations <- list()

for (city in city_names) {
  city_data <- city_totals %>%
    filter(cook_municipality_name == city) %>%
    arrange(year)

  city_tifs <- tif_creation_by_city %>%
    filter(cook_municipality_name == city) %>%
    arrange(tif_creation_year) %>%
    mutate(label_y_offset = row_number())

  earliest          <- min(city_tifs$tif_creation_year, na.rm = TRUE)
  city_data_filtered <- city_data %>% filter(year >= earliest)
  first_frozen      <- city_data_filtered %>%
    filter(!is.na(frozen_eav) & frozen_eav > 0) %>%
    pull(frozen_eav) %>%
    first()

  max_eav    <- max(city_data$total_eav, na.rm = TRUE)
  city_label <- gsub("VILLAGE OF ", "", city)

  muni_plot <- muni_plot %>% add_lines(
    x         = city_data$year,
    y         = city_data$total_eav,
    fill      = "tozeroy",
    fillcolor = "rgba(128, 128, 128, 0.3)",
    line      = list(color = "grey"),
    name      = "Total EAV",
    visible   = FALSE,
    hoverinfo = "text",
    text      = paste0(city_label, " Total EAV: $", round(city_data$total_eav / 1e6, 2), "M")
  )

  muni_plot <- muni_plot %>% add_polygons(
    x         = c(city_data_filtered$year, rev(city_data_filtered$year)),
    y         = c(city_data_filtered$total_eav, rep(first_frozen, nrow(city_data_filtered))),
    fillcolor = "rgba(145, 61, 136, 0.5)",
    line      = list(width = 0),
    visible   = FALSE,
    hoverinfo = "skip"
  )

  for (i in 1:nrow(city_tifs)) {
    muni_plot <- muni_plot %>% add_segments(
      x        = city_tifs$tif_creation_year[i],
      xend     = city_tifs$tif_creation_year[i],
      y        = 0,
      yend     = max_eav,
      line     = list(color = "black", dash = "dash", width = 1),
      visible  = FALSE,
      hoverinfo = "none"
    )
  }

  city_annos <- lapply(1:nrow(city_tifs), function(i) {
    list(
      x         = city_tifs$tif_creation_year[i],
      y         = max_eav * (0.95 - 0.08 * (i - 1)),
      text      = paste0(
        stringr::str_trunc(city_tifs$tax_tif_district_name[i], 35),
        "\n", city_tifs$tif_creation_year[i]
      ),
      showarrow = FALSE,
      xanchor   = "left",
      font      = list(size = 9, color = "black")
    )
  })

  all_city_annotations[[city]] <- city_annos
  traces_per_city_v2 <- c(traces_per_city_v2, 2 + nrow(city_tifs))
}

cumulative_traces_v2 <- c(0, cumsum(traces_per_city_v2))

buttons_v2 <- lapply(seq_along(city_names), function(i) {
  vis <- rep(FALSE, sum(traces_per_city_v2))
  vis[(cumulative_traces_v2[i] + 1):cumulative_traces_v2[i + 1]] <- TRUE
  list(
    method = "update",
    args   = list(
      list(visible = vis),
      list(
        annotations = all_city_annotations[[city_names[i]]],
        title       = gsub("VILLAGE OF ", "", city_names[i])
      )
    ),
    label  = gsub("VILLAGE OF ", "", city_names[i])
  )
})

muni_plot <- muni_plot %>% layout(
  xaxis       = list(title = "Year", showgrid = FALSE),
  yaxis       = list(title = "Total EAV", tickformat = "$~s"),
  showlegend  = FALSE,
  updatemenus = list(list(
    active  = 0,
    buttons = buttons_v2,
    x       = 0.1,
    y       = 1.2
  )),
  annotations = all_city_annotations[[city_names[1]]],
  margin      = list(t = 100)
)

muni_plot <- plotly_build(muni_plot)
for (j in 1:traces_per_city_v2[1]) {
  muni_plot$x$data[[j]]$visible <- TRUE
}
muni_plot
```

## City Level Analysis of Frozen vs TIF EAV

### Phoenix

The Village of Phoenix represents the most extreme concentration of TIFs, with 98% of all PINs located within TIF districts. With a stable frozen base of \$7.2 million, the increment started at \$9.8 million in 2006 and peaked at \$11.9 million in 2008. From there, total EAV declined steadily through the mid-2010s, compressing the increment to a low of \$4.9 million in 2015. A modest recovery followed, but total EAV has remained range-bound between \$12--14 million through most of the period. By 2023, total EAV recovered to \$19.6 million, yielding an increment of \$12.4 million comparable to its 2006 level, suggesting limited net growth in taxable value over nearly two decades.

### Bedford Park

The Village of Bedford Park shows one of the most dramatic increment trajectories. With a frozen base anchored at \$12.1 million, the increment started large with approximately \$110.7 million in 2008, reflecting a high pre-existing property value base at TIF creation. The increment then declined sharply as total EAV fell through the early 2010s, reaching a low of \$58.6 million in 2015. From 2016 onward, total EAV recovered consistently, reaching \$197.9 million by 2023 and generating an increment of \$185.8 million — the largest among all five cities and nearly double its 2008 peak.

### Bellwood

The Village of Bellwood follows a similar pattern to Bedford Park, but with an even lower frozen base of \$2.7 million. As a result, nearly the entire total EAV appears as increment on the graph. The increment peaked at \$187.3 million in 2008 before declining through the early 2010s to a low of \$107 million in 2015. Recovery was gradual but consistent, and by 2023 total EAV reached \$217.4 million, generating an increment of \$214.7 million — the highest in absolute terms across all five cities. The low frozen base means Bellwood's TIF captures almost all property value growth within its boundaries.

### Ford Heights

The Village of Ford Heights has the most modest TIF activity of the five cities. The TIF was not established until 2015, so the graph shows no increment area before that year. With a frozen base of \$2.3 million, the increment has remained small throughout, peaking at \$2.2 million in 2020 when total EAV reached \$4.5 million. Since then, both total EAV and the increment have declined, falling to \$3.2 million and \$900,000 respectively by 2023. Ford Heights illustrates a case where TIF designation has not driven significant property value growth.

### Posen

The Village of Posen presents the most stable increment trajectory among the five cities. With a frozen base of \$11.8 million in place since 2006, the increment has remained consistently positive throughout the entire period. It peaked at \$15.1 million in 2008, declined gradually to a trough of \$5.5 million in 2015, then recovered steadily. A notable jump in 2020 pushed total EAV to \$28 million, and by 2023 it reached \$30.6 million, yielding an increment of \$18.8 million. Unlike the other cities, Posen's increment area shows a smooth, stable band with no sharp contractions, reflecting consistent property value retention across the TIF period.

# Part III: Tax Rate Impact of TIF Districts

### What is effective TAX rate?

The effective tax rate is the rate that taxing agencies like schools, parks, the municipality itself must levy to collect the revenue they need. It is calculated as ([Source](https://tax.illinois.gov/content/dam/soi/en/web/tax/localgovernments/property/documents/ptax-1-a.pdf)):

$$\text{Effective Rate} = \frac{\text{Levy}}{\text{Taxable EAV Base}}$$

TIF districts shrink the taxable base available to these agencies. When a TIF is created, the increment is diverted into the TIF fund and removed from the general levy base. Agencies still need to collect the same revenue, so they apply their levy to a smaller base, which pushes the rate up.

This matters especially in our five cities because the proportion of TIF'd PINs is high. Phoenix (98%) and Bellwood have very little taxable base left outside their TIF boundaries. The question is: **by how much are tax rates inflated as a result?**

With TIF means the rate agencies actually charge today, levy divided by the frozen base only. Without TIF is the counterfactual: if the increment had never been diverted, the base would be larger and the rate lower. The difference between the two is the rate inflation attributable to the TIF.

## Temporal Effective Tax Rate Analysis

The chart below tracks this rate inflation from 2006 to 2023. Solid lines show the actual rate agencies charged and dashed lines show what the rate would have been if TIF increment had remained in the taxable base.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Build agency_nums_2023: unique tax code / agency combinations for target cities
# Coerce to character before joining to avoid type mismatch during knit
agency_nums_2023 <- tif_dists %>%
  filter(year == 2023) %>%
  mutate(tax_code_num = as.character(tax_code_num)) %>%
  select(tax_code_num, agency_num) %>%
  distinct() %>%
  left_join(
    target_cities %>%
      mutate(tax_code = as.character(tax_code)) %>%
      select(tax_code, cook_municipality_name) %>%
      distinct(),
    by = c("tax_code_num" = "tax_code")
  ) %>%
  filter(!is.na(cook_municipality_name))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# 1. Get agencies for all years
agency_details_all <- lookup_agency(
  year     = 2006:2023,
  tax_code = unique(agency_nums_2023$tax_code_num)
)

# 2. Attach agency details
agency_city_all <- agency_details_all %>%
  mutate(tax_code = as.character(tax_code)) %>%
  left_join(
    agency_nums_2023 %>%
      select(tax_code_num, cook_municipality_name) %>%
      mutate(tax_code_num = as.character(tax_code_num)) %>%
      distinct(),
    by = c("tax_code" = "tax_code_num")
  ) %>%
  filter(!is.na(cook_municipality_name), !is.na(year))

# 3. TIF increment for all years
increment_all <- city_totals %>%
  select(cook_municipality_name, year, amt_to_tif)

# 4. Calculate rates for all years
rate_impact_all <- agency_city_all %>%
  left_join(increment_all, by = c("cook_municipality_name", "year")) %>%
  group_by(year, cook_municipality_name, agency_num, agency_name) %>%
  summarize(
    levy       = first(agency_total_ext),
    eav_actual = first(agency_total_eav),
    amt_to_tif = first(amt_to_tif),
    .groups    = "drop"
  ) %>%
  mutate(
    eav_counterfactual  = eav_actual + amt_to_tif,
    rate_actual         = levy / eav_actual,
    rate_counterfactual = levy / eav_counterfactual
  )

# 5. City-year summary
city_rate_temporal <- rate_impact_all %>%
  group_by(year, cook_municipality_name) %>%
  summarize(
    rate_with_tif    = round(mean(rate_actual, na.rm = TRUE) * 100, 3),
    rate_without_tif = round(mean(rate_counterfactual, na.rm = TRUE) * 100, 3),
    .groups          = "drop"
  ) %>%
  mutate(city_label = gsub("VILLAGE OF ", "", cook_municipality_name))

# 6. Colors
city_colors <- setNames(
  RColorBrewer::brewer.pal(5, "Set1"),
  unique(city_rate_temporal$city_label)
)

# 7. Plot
p_rates <- plot_ly()

for (city in unique(city_rate_temporal$city_label)) {
  cd  <- city_rate_temporal %>% filter(city_label == city)
  col <- city_colors[[city]]

  p_rates <- p_rates %>% add_lines(
    data        = cd,
    x           = ~year,
    y           = ~rate_with_tif,
    name        = city,
    legendgroup = city,
    showlegend  = TRUE,
    line        = list(color = col, width = 2, dash = "solid")
  )

  p_rates <- p_rates %>% add_lines(
    data        = cd,
    x           = ~year,
    y           = ~rate_without_tif,
    name        = paste0(city, " (w/o TIF)"),
    legendgroup = city,
    showlegend  = TRUE,
    line        = list(color = col, width = 2, dash = "dash")
  )
}

p_rates %>% layout(
  title     = list(
    text = "Effective Tax Rate With vs. Without TIF Increment (2006–2023)",
    x    = 0.5
  ),
  xaxis     = list(title = "Year", dtick = 2),
  yaxis     = list(title = "Average Effective Tax Rate (%)", ticksuffix = "%"),
  legend    = list(orientation = "v", x = 1.02, y = 1),
  hovermode = "x unified",
  margin    = list(r = 160)
)
```

The gap between solid and dashed lines varies considerably across cities and widens over time in Phoenix as their increments grew, suggesting the rate inflation burden intensified as those TIFs matured.

### Findings

Effective tax rates rose across all five cities between 2006 and 2023, a trend potentially driven largely by the post-2008 property value collapse compressing the taxable base countywide, forcing agencies to raise rates to meet their levies. TIF base compression shows the burden TIFs can be to tax rate increases, most severely in Phoenix, where nearly all taxable property sits inside a TIF district and has for nearly two decades.

-   Ford Heights carried the highest absolute rates throughout, reaching 3.0% in 2014, reflecting its broader fiscal distress rather than TIF activity specifically. Its solid and dashed lines track nearly identically until the TIF is established in 2015.
-   Bedford Park and Bellwood show a different pattern: their lines overlap almost perfectly through 2015, then diverge sharply as their large increments matured. By 2023 Bedford Park's actual rate is 1.02% against a counterfactual of 0.88%, and Bellwood's is 1.22% against 1.02%.
-   Posen shows a modest but stable gap throughout.
-   Phoenix stands out the most, where the gap between the actual rate and the counterfactual rate without TIF increment has been large and persistent since 2006, and widens steadily over time. By 2023 the actual rate of 2.05% is nearly 46% higher than the 1.40% counterfactual, the largest TIF-induced inflation of any city in the dataset.

### Phoenix: The Outlier

The apparent drop in Phoenix's effective tax rate in 2023 is not a data anomaly but rather reflects Cook County's triennial reassessment cycle. The south suburbs, where Phoenix is located, underwent reassessment in 2023, which pushed Phoenix's total EAV from approximately \$13 million in 2022 to \$19.6 million which is a 50% increase in a single year. This expanded base temporarily compressed the effective rate, producing the visible dip in the chart. The underlying structural story, however, remains unchanged: the Village of Phoenix levies against a frozen TIF base of roughly \$5.4 million while its actual total EAV is nearly four times that figure, with the difference captured entirely by the TIF increment. This is not a transient measurement artifact but a persistent fiscal condition. Phoenix has operated with nearly its entire property tax base locked inside TIF boundaries since at least 2006, leaving the municipality, park district, and library perpetually levying against a fraction of the value that actually exists within their borders. The reassessment year simply provides a moment of temporary relief within an otherwise structural constraint.

## Agency-Level Breakdown: Who Bears the Burden?

The city-level summary tells us the aggregate rate inflation, but it masks which taxing agencies are most exposed. A school district, a park district, and the municipality itself all levy against the same frozen base, but their sensitivity to that compression differs based on how large their levy is and how geographically concentrated their base is.

This matters because the fiscal consequences of TIFs are not borne equally. In Illinois, school districts are typically the largest single claimant of property tax revenue. When TIF increment is withheld from the general base, schools effectively subsidize TIF-funded development whether or not they are party to the redevelopment agreement. In distressed communities like Ford Heights and Phoenix, where school funding is already strained, this compression is worth examining directly. ([Source](https://www.civicfed.org/blog/setting-record-straight-how-tax-increment-financing-tif-actually-interacts-chicago-public))

### TIF-Induced Rate Inflation Over Time

The chart below shows rate inflation by agency type for each city from 2006 to 2023. Use the dropdown to switch cities and single-click to hide a legend item or double-click to isolate a single agency type.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rate_by_type <- agency_city_all %>%
  left_join(increment_all, by = c("cook_municipality_name", "year")) %>%
  filter(!grepl("TIF", agency_name, ignore.case = TRUE)) %>%
  group_by(year, cook_municipality_name, agency_minor_type) %>%
  summarize(
    levy                = sum(agency_total_ext, na.rm = TRUE),
    eav_actual          = first(agency_total_eav),
    amt_to_tif          = first(amt_to_tif),
    .groups             = "drop"
  ) %>%
  mutate(
    eav_counterfactual  = eav_actual + amt_to_tif,
    rate_actual         = levy / eav_actual,
    rate_counterfactual = levy / eav_counterfactual,
    rate_pct_inflation  = ((rate_actual - rate_counterfactual) / rate_counterfactual) * 100,
    city_label          = gsub("VILLAGE OF ", "", cook_municipality_name)
  )

agency_types <- unique(rate_by_type$agency_minor_type)
type_colors  <- setNames(
  colorRampPalette(brewer.pal(8, "Set2"))(length(agency_types)),
  agency_types
)

p_type <- plot_ly()

for (city in unique(rate_by_type$city_label)) {
  for (atype in agency_types) {
    cd <- rate_by_type %>%
      filter(city_label == city, agency_minor_type == atype)
    if (nrow(cd) == 0) next

    p_type <- p_type %>% add_lines(
      data        = cd,
      x           = ~year,
      y           = ~rate_pct_inflation,
      name        = atype,
      legendgroup = atype,
      showlegend  = TRUE,
      visible     = (city == unique(rate_by_type$city_label)[1]),
      line        = list(color = type_colors[[atype]], width = 2)
    )
  }
}

traces_per_city_type <- rate_by_type %>%
  group_by(city_label) %>%
  summarize(n_types = n_distinct(agency_minor_type), .groups = "drop") %>%
  pull(n_types)

cum_type        <- c(0, cumsum(traces_per_city_type))
total_type      <- sum(traces_per_city_type)
city_labels_type <- unique(rate_by_type$city_label)

buttons_type <- lapply(seq_along(city_labels_type), function(i) {
  vis <- rep(FALSE, total_type)
  vis[(cum_type[i] + 1):cum_type[i + 1]] <- TRUE
  list(
    method = "update",
    label  = city_labels_type[i],
    args   = list(
      list(visible = vis),
      list(title = list(
        text = paste0(city_labels_type[i], ": Rate Inflation by Agency Type"),
        x    = 0.5
      ))
    )
  )
})

p_type %>% layout(
  title = list(
    text = paste0(city_labels_type[1], ": Rate Inflation by Agency Type"),
    x    = 0.5
  ),
  xaxis       = list(title = "Year", dtick = 2),
  yaxis       = list(title = "Rate Inflation (%)", ticksuffix = "%"),
  updatemenus = list(list(
    active      = 0,
    buttons     = buttons_type,
    x           = 0.01,
    xanchor     = "left",
    y           = 1.12,
    yanchor     = "top",
    bgcolor     = "white",
    bordercolor = "#aaa",
    font        = list(size = 12)
  )),
  hovermode = "x unified",
  legend    = list(
    orientation     = "v",
    x               = 1.02,
    y               = 1,
    itemdoubleclick = "toggleothers"
  ),
  margin = list(t = 90, r = 160)
)
```

### Findings

The agency-type breakdown reveals that the timing and magnitude of TIF-induced rate inflation varies sharply across cities, but the pattern of which agencies bear the burden is consistent.

-   In Bedford Park, Bellwood, and Ford Heights, rate inflation was effectively zero before 2014–2015 and only began climbing as TIF increments matured. By 2023 Bedford Park's elementary school district faces nearly 20% rate inflation and Bellwood's municipal agency exceeds 35%, driven by the large increments that accumulated in those districts over the prior decade.
-   Posen shows a more moderate and steady pattern, with park and municipal agencies carrying 15–27% inflation throughout the period, reflecting its older and more stable TIF base.
-   Phoenix is the outlier in both scale and structure. Its park and municipal agencies have faced rate inflation between 100–200% for nearly the entire period since 2006, a direct consequence of having almost its entire tax base locked inside TIF boundaries from the outset, leaving those agencies levying against a frozen base that represents a small fraction of actual property value.

Across all five cities, park districts and municipalities consistently show the highest rate inflation, while county-level and school agencies follow with comparatively lower figures.

# Conclusion

This report examined TIF districts across five south suburban Cook County municipalities: Phoenix, Bedford Park, Bellwood, Ford Heights, and Posen. Each selected for having among the highest share of TIF-ed parcels outside the City of Chicago. Across all five, TIF districts capture a substantial portion of total property value as increment, ranging from modest levels in Ford Heights to nearly the entirety of taxable EAV in Phoenix. The frozen base, by design, remains fixed at the time of TIF creation, meaning that decades of property value growth flow into TIF funds rather than the general tax base available to schools, parks, and municipalities.

The tax rate analysis makes the fiscal consequence concrete. Agencies levying against the frozen base must charge higher rates to meet the same levy, and the gap between actual and counterfactual rates widens as increments grow. Phoenix carries the most severe distortion, with actual rates running nearly 46% above the counterfactual by 2023, a gap that has persisted since at least 2006. Bedford Park and Bellwood show similar divergence emerging after 2015 as their large industrial increments matured. Park districts and municipal agencies consistently bear the greatest rate inflation across all five cities.

Ford Heights offers a useful contrast. Its TIF was created in 2015 and has generated minimal increment, with total EAV in the district remaining well below levels seen in the other four cities. The near-identical actual and counterfactual rate lines through most of the period confirm that TIF-induced rate inflation is directly proportional to increment size, indicating where increments are small, the fiscal burden on other agencies remains limited.
